\vspace{-15mm}\fontsize{16pt}{10pt}\selectfont\textbf{Facial Expression Recognition with\\ Convolutional Neural Networks}

### 2018-04-12 doc/paper.tex
references and discussion

\usepackage{booktabs} % Horizontal rules in tables
\usepackage{float} % Required for tables and figures in the multi-column environment - they need to be placed in specific locations with the [H] (e.g. \begin{table}[H])
\usepackage{hyperref} % For hyperlinks in the PDF

@@

Automatically recognizing facial expressions is an interesting and challenging problem in many fields, including human computer interaction and artificial emotional intelligence. Facial expressions can be interpreted and categorized into six classes of basic emotion that seem to be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise [REF]. This projects goal is to develop a classifier for real-time detection of these facial expressions.

@@

\section{Related work}
In the last years, great progress was made on the problem of automatically detecting facial expression from images. Some researchers have attempted to derive emotions from detecting individual muscle movements in images of faces by identifying muscles and their degree of activation [REF]. SIFT aggregation [REF]… Manual feature extraction tends to require expert knowledge and extra work for implementing complex algorithms, so it is convenient to implicitly extract relevant features from the data. This can be accomplished with Convolutional Neural Networks (CNNs) that are trained on a set of labeled images.

@@

\section{Datasets}
To train an automatic classifier using deep learning technology, a dataset that describes the problem is required. Each image in the dataset is mapped to a matching integer class label for the corresponding emotion. Two datasets, FER2013 [REF] and CK+ [REF] have been considered for training the deep learning model.

@@

The class labels were crowd-sourced with Amazon Mechanical Turk, where quality work is not incentivized, so we can expect noisy labels to be a problem. It is expected that every subject shows only a single emotion, where in reality faces can express mixed emotions (e.g happiness and surprise). A research group at Microsoft (who?) decided to minimize label-noise in FER-2013 by re-labeling the dataset (REF). Ten workers on Mechanical Turk were asked to label each image, yielding a probability distribution of emotion classes instead of a single class for each image.

\subsection{Cohn-Kanade (CK+)}
To augument the wild variety of FER-2013 images, a second, more conservative dataset with only frontal faces is considered. The Cohn-Kanade (CK+) dataset consists of 593 image sequences across 123 subjects. Each image sequence goes from a neutral facial expression to a "peak expression" according to the six basic emotions. Only 327 of the 593 image sequences are labeled with emotion data. This is because not all images fit the prototypic definition of the Facial Action Coding System (FACS) according to Ekman et al. CK+ also contains FACS data, in which facial action units (e.g raised eyebrow) are identified along with their degree of activation. To show that extracted features are not strictly required for training an automatic classifier, the FACS data is not used as input.

@@

\subsubsection{Convolutional Layer}
The convolutional layer is the core building block of a CNN. Each layer is initialized with a set of $n$ random two-dimensional filter matrices (eg 3x3 pixels) and a pixel-stride. When an image is put into the convolutional layer, each of the filters is moved across the height and width of the image, calculating a dot product between the filter matrix and the current section of the image. The resulting matrices of dot products are called feature maps. To introduce non-linearity to the operation, a ReLu activation function $f(n) = max(0, n)$ is used on the feature-maps element-wise. The filters in the convolutional layers are trained using a backpropagation algorithm (REF). Intuitively, filters activate when they see different types of visual features, such as vertical or diagonal edges on the lower layers or, in the top convolutional layers, more complex patterns like eyebrows or mouths.

@@

\subsection{AlexNet-like CNN}
For the Imagenet Challenge (REF) Alex … proposed a deep neural network (DNN) model consisting of alternating convolution and pooling layers with local response normalization in between, followed by three fully connected layers.

@@

% - CNN robustness to occlusion, distortions and geometric transformations.

@@

\begin{thebibliography}{9}

	% OpenCV Haarcascade Classifier
	\bibitem{violajones}
  	P. A. Viola and M. J. Jones. Robust real-time face detection.
	International Journal of Computer Vision
	57(2):137–154, 2004.

	% the ICML 2013 workshop’s facial expression recognition challenge
	\bibitem{replearning}
	I. J. Goodfellow et al.
	Challenges in representation learning: A report on three machine
	learning contests.
	Neural Networks
	, 64:59 – 63, 2015.
	Special Issue on ‘Deep Learning of Representations’

	%[LeCun and Bengio, 1995a]
    %Y. LeCun and Y. Bengio. Convolutional networks for images, speech, and time-series. 
    %In M. A. Arbib, 		editor, The Handbook of Brain Theory and Neural Networks. MIT Press, 1995

    % Paul Ekmans 1969 research paper "Facial Expressions"

\end{thebibliography}

>>>>

\usepackage{booktabs} % Horizontal rules in tables
\usepackage{float} % Required for tables and figures in the multi-column environment - they need to be placed in specific locations with the [H] (e.g. \begin{table}[H])
\usepackage{hyperref} % For hyperlinks in the PDF
\usepackage{cite} % BibTex

@@

Automatically recognizing facial expressions is an interesting and challenging problem in many fields, including human computer interaction and artificial emotional intelligence. Facial expressions can be interpreted and categorized into six classes of basic emotion that seem to be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise \cite{ekman93}. This projects goal is to develop a classifier for real-time detection of these facial expressions.

@@

\section{Related work}
In the last years, great progress was made on the problem of automatically detecting facial expression from images. Some researchers have attempted to derive emotions from detecting individual muscle movements in images of faces by identifying muscles and their degree of activation \cite{lien98}. Manual feature extraction tends to require expert knowledge and extra work for implementing complex algorithms, so it is convenient to implicitly extract relevant features from the data. This can be accomplished with Convolutional Neural Networks (CNNs) that are trained on a set of labeled images.

@@

\section{Datasets}
To train an automatic classifier using deep learning technology, a dataset that describes the problem is required. Each image in the dataset is mapped to a matching integer class label for the corresponding emotion. Two datasets, FER2013 \cite{goodfel13} and CK+ \cite{cohn00} have been considered for training the deep learning model.

@@

The class labels were crowd-sourced with Amazon Mechanical Turk, where quality work is not incentivized, so we can expect noisy labels to be a problem. It is expected that every subject shows only a single emotion, where in reality faces can express mixed emotions (e.g happiness and surprise). A research group at Microsoft decided to minimize label-noise in FER-2013 by re-labeling the dataset \cite{barsoum16}. Ten workers on Mechanical Turk were asked to label each image, yielding a probability distribution of emotion classes instead of a single class for each image.

\subsection{Cohn-Kanade (CK+)}
To augment the wild variety of FER-2013 images, a second, more conservative dataset with only frontal faces is considered. The Cohn-Kanade (CK+) dataset consists of 593 image sequences across 123 subjects. Each image sequence goes from a neutral facial expression to a "peak expression" according to the six basic emotions. Only 327 of the 593 image sequences are labeled with emotion data. This is because not all images fit the prototypic definition of the Facial Action Coding System (FACS) according to Ekman et al. CK+ also contains FACS data, in which facial action units (e.g raised eyebrow) are identified along with their degree of activation. To show that extracted features are not strictly required for training an automatic classifier, the FACS data is not used as input.

@@

\subsubsection{Convolutional Layer}
The convolutional layer is the core building block of a CNN. Each layer is initialized with a set of $n$ random two-dimensional filter matrices (eg 3x3 pixels) and a pixel-stride. When an image is put into the convolutional layer, each of the filters is moved across the height and width of the image, calculating a dot product between the filter matrix and the current section of the image. The resulting matrices of dot products are called feature maps. To introduce non-linearity to the operation, a ReLu activation function $f(n) = max(0, n)$ is used on the feature-maps element-wise. The filters in the convolutional layers are trained using a backpropagation algorithm. Intuitively, filters activate when they see different types of visual features, such as vertical or diagonal edges on the lower layers or, in the top convolutional layers, more complex patterns like eyebrows or mouths.

@@

%\subsection{AlexNet-like CNN}
%For the Imagenet Challenge (REF) Alex … proposed a deep neural network (DNN) model consisting of alternating convolution and pooling layers with local response normalization in between, followed by three fully connected layers.

@@

% - CNN robustness to occlusion, distortions and geometric transformations.
The final model scored an accuracy of ~75\% on a test-set of 3557 samples. In live predictions with a webcam, the classifier has shown to be a cool gadget to play around with. However, some emotions such as disgust and fear are hard to reproduce predictions for. This is most likely because the seven classes of emotion in the dataset are highly imbalanced. For instance disgust only has 500 samples while happy has much more. This results in biased predictions with higher probabilities of the majority classes. In order to balance the training dataset, data augmentation could be used: To increase the number of training examples for a given class, the images could be flipped horizontally or rotated in multiple random angles. For normalizing the images, it is a common practice to subtract the mean pixel intensities over the whole dataset from each picture. Given more time for this project, I would like to have experimented with such techniques of data augmentation.

@@

%\begin{thebibliography}{9}
%
%	% OpenCV Haarcascade Classifier
%	\bibitem{violajones}
%  	P. A. Viola and M. J. Jones. Robust real-time face detection.
%	International Journal of Computer Vision
%	57(2):137–154, 2004.
%
%	% the ICML 2013 workshop’s facial expression recognition challenge
%	\bibitem{replearning}
%	I. J. Goodfellow et al.
%	Challenges in representation learning: A report on three machine
%	learning contests.
%	Neural Networks
%	, 64:59 – 63, 2015.
%	Special Issue on ‘Deep Learning of Representations’
%	
%	%[LeCun and Bengio, 1995a]
%    %Y. LeCun and Y. Bengio. Convolutional networks for images, speech, and time-series. 
%    %In M. A. Arbib, 		editor, The Handbook of Brain Theory and Neural Networks. MIT Press, 1995
%    
%    % Paul Ekmans 1969 research paper "Facial Expressions"
%
% 
%\end{thebibliography}

\bibliography{paper}
\bibliographystyle{plain}
